Automatically Proving Equivalence by Type-Safe Reflection

### 2017-05-20 paper-CICM-17/content/1-introduction.tex
Add a bit more stuff, prune the bibliography

Proofs assistants like Coq~\cite{BertotC04} 
and programming languages like 
Agda~\cite{DBLP:conf/tldi/Norell09} and
Idris~\cite{brady2013idris} are based on Intensional Type Theories that contain
two notions of equality: propositional equality, that can be
manipulated in the language, and judgemental (or definitional) equality.
Propositional equality corresponds to the mathematical notion: this is
a proposition that can be assumed, negated, proved or disproved. Since in type
theory, propositions are types~\cite{How80}, the proposition that two
elements $x$ and $y$ are equal corresponds to a type.
If $x$ and $y$ are of type $a$, then the type
$Id_a(x, y)$ represents the proposition ``$x$ is equal to $y$". If this type is
inhabited, then $x$ is said to be provably equal to $y$. Thus, $Id$ is a type family (parameterised by the type $a$) indexed over two elements of $a$, giving $\code{Id}\ (a:\code{Type}):a \rightarrow a \rightarrow \code{Type}$. For
convenience, we write $(\code{Id}_a\ x\ y)$ as $(x\ =_a\ y)$.

@@

Proving that one term is equal to another is common in formal verification, and
proof obligations arise naturally in dependently typed programming when
indexing types over values in order to capture some logical properties.
To demonstrate this,
we revisit an example from previous work~\cite{DBLP:conf/plpv/Brady13} which
shows how proof obligations arise when a type is indexed by natural
numbers. Our goal is to implement a verified
library of binary numbers. To ensure functional correctness,
we define the types \code{Bit} and \code{Binary} indexed over the value they represent (expressed as a natural number):

@@

We will write a function to add two binary numbers.
To do so, we begin with an auxiliary function, which adds three bits
(the third is a carry bit), and produces the two bits of
the result, where the first is the most significant bit:

@@

\codeL{((c\ + (bit2\ + (v1\ + (v1\ + 0)))) + (bit\ + (v\ + (v\ + 0))))}
while we're trying to provide a term indexed over

@@

The definition of \code{adc} we have given would behave correctly, and it has
\emph{provably} the expected type, but it does not have it \emph{immediately}
or \emph{judgementally}: after full reductions the expected and provided types are still
different.
If we use provisional definitions~\cite{DBLP:conf/plpv/Brady13}, the previous
definition can be accepted by the system, but we now have two proof obligations
\code{adc\_lemma\_1} and \code{adc\_lemma\_2} demanding proofs of equality
between the two types, and they have to be provided by hand.
For example, using a proof script:

@@

Our handwritten proof \code{adc\_lemma\_2} uses only the existence of a neutral
element, and the associativity and commutativity of $+$ on \code{Nat}. Thus,
we're rewriting a term by using the properties of a commutative monoid. We
would like a generic prover for commutative monoids to find a proof 
automatically.

>>>>

Proofs assistants like Coq~\cite{BertotC04} 
and programming languages like 
Agda~\cite{norell2007thesis} and
Idris~\cite{brady2013idris} are based on Intensional Type Theories that contain
two notions of equality: propositional equality, that can be
manipulated in the language, and judgemental (or definitional) equality.
Propositional equality corresponds to the mathematical notion: this is
a proposition that can be assumed, negated, proved or disproved. Since in type
theory, propositions are types~\cite{How80}, the proposition that two
elements $x$ and $y$ are equal corresponds to a type.
If $x$ and $y$ are of type $a$, then the type
$Id_a(x, y)$ represents the proposition ``$x$ is equal to $y$". If this type is
inhabited, then $x$ is said to be provably equal to $y$. Thus, $Id$ is a type family (parameterised by the type $a$) indexed over two elements of $a$, giving $\code{Id}\ (a:\code{Type}):a \rightarrow a \rightarrow \code{Type}$. For
convenience, we write $(\code{Id}_a\ x\ y)$ as $(x\ =_a\ y)$.

@@

Proving that one term is equal to another is common in formal verification, and
proof obligations arise naturally in dependently typed programming when
indexing types over values in order to capture some logical properties.
To demonstrate this,
we revisit an example from previous work~\cite{BradyTFP07} which
shows how proof obligations arise when a type is indexed by natural
numbers. Our goal is to implement a verified
library of binary numbers. To ensure functional correctness,
we define the types \code{Bit} and \code{Binary} indexed over the value they represent (expressed as a natural number):

@@

We will write a function to add two binary numbers.
To do so, we begin with an auxiliary function, which adds three bits
(the third is a carry bit), and produces the two bits of
the result, where the first is the more significant bit:

@@

\codeL{((c\ + (bit2\ + (v1\ + (v1\ + 0)))) + (bit\ + (v\ + (v\ + 0))))}
\\
while we're trying to provide a term indexed over:

@@

The definition of \code{adc} we have given would behave correctly, and it has
\emph{provably} the expected type, but it does not have it \emph{immediately}
or \emph{judgementally}: after full reductions the expected and provided types are still
different.
To make the previous
definition acceptable, we need to solve two proof obligations
\code{adc\_lemma\_1} and \code{adc\_lemma\_2} which demand proofs of equality
between the two types.
For example, using a proof script:

@@

Our handwritten proof \code{adc\_lemma\_2} uses only the existence of a neutral
element, and the associativity and commutativity of $+$ on \code{Nat}. Thus,
we're rewriting a term by using the properties of a commutative monoid. 
With the right choice of combinators, such proofs could be made much simpler,
but we would like a generic prover for commutative monoids to find a proof
automatically.

### 2017-05-20 paper-CICM-17/content/3-hierarchyOfTactics.tex
Add a bit more stuff, prune the bibliography

For an expression $e_x\ :\ \code{ExprMa}\ \Gamma\ x$, we say that $e_x$
denotes the value $x$ in the context $\Gamma$.  When an expression
is a variable \code{(VarMa\ \_\ $\Gamma$\ i)}, the denoted value is the
corresponding variable in the context, i.e., \code{(index\ i\ $\Gamma$)}.  The
expression \code{(ConstMa\ \_\ $\Gamma$\ k)} denotes the constant $k$ in any context $\Gamma$. Finally, if $e_x$ is an expression encoding the concrete value $x$, and
$e_y$ is an expression encoding the concrete value $y$, then the expression
\code{PlusMa\ $e_x$\ $e_y$} denotes the concrete value $(x + y)$. Because the
reflected terms embed their corresponding inputs, they are guaranteed to be
sound representations.

@@

We have built an automatic reflection mechanism which enables the machine to
build the reflected terms for us automatically. This is not essential to our
approach, but it simplifies the usage of the tactics by removing the need to
write long encodings by hand.  To do so, we used Idris' reflection mechanism,
which enables pattern matching on syntax, rather than on constructors.

>>>>

For an expression $e_x\ :\ \code{ExprMa}\ \Gamma\ x$, we say that $e_x$
denotes the value $x$ in the context $\Gamma$.  When an expression
is a variable \code{(VarMa\ \_\ $\Gamma$\ i)}, the denoted value is the
corresponding variable in the context, i.e., \code{(index\ i\ $\Gamma$)}.  The
expression \code{(ConstMa\ \_\ $\Gamma$\ k)} denotes the constant $k$ in any context $\Gamma$. Finally, if $e_x$ is an expression encoding the concrete value $x$, and
$e_y$ is an expression encoding the concrete value $y$, then the expression
\code{PlusMa\ $e_x$\ $e_y$} denotes the concrete value $(x + y)$. Because the
reflected terms embed their corresponding inputs, they are guaranteed to be
sound representations.
This is a \emph{local} approach to syntax representation~\cite{Farmer13}
in that the reflected representation will \emph{only} represent terms in
a magma.

@@

We have built an automatic reflection mechanism which enables the machine to
build the reflected terms for us automatically. This is not essential to our
approach, but it simplifies the usage of the tactics by removing the need to
write long encodings by hand.  To do so, we used Idris' reflection mechanism,
which enables pattern matching on syntax, rather than on constructors. 
While we omit the full details due to space constraints, reflecting values
involves defining functions of the following form:

\begin{lstlisting}
%reflection reflectGroup : (p : Group c) $\rightarrow$ ($\Gamma$ : Vect n c) $\rightarrow$ 
                           (x : c) $\rightarrow$ ($\Gamma'$ ** ExprG p ($\Gamma'$ ++ $\Gamma$) x)
\end{lstlisting}

The \texttt{\%reflection} annotation means that Idris treats this as a 
compile time function on \emph{syntax}. Given a value of type
\texttt{c}, in some context $\Gamma$, it constructs a reflected expression 
in an extended context \texttt{$\Gamma'$ ++ $\Gamma$}. The context
contains references to subexpressions 
which are not themselves representable by \texttt{ExprG}.

### 2017-05-20 paper-CICM-17/content/4-relatedWork.tex
minor changes before Edwin shortens the paper
Add a bit more stuff, prune the bibliography

As mentioned, Coq's implementation of a ring prover~\cite{Coq2005} was also using proof-by-reflection techniques, but without the guarantees obtained with our type-safe reflection mechanism, and without the correct-by-construction approach : they have first defined the normalisation of terms, and afterwards they've proven the correctness of the normalisation with the external lemma :
$\forall\ (e1\ e2\ :\ \code{Expr}),\ beq_{Expr}\ (\code{norm}\ e1)\ (\code{norm}\ e2)\ =\ true\ \rightarrow\ \code{reify}\ e1\ \simeq\ \code{reify}\ e2$ which therefore needs a \code{reify} function that we do not need in our development. Another difference is that Coq's prover only deals with rings and semi-rings (commutative or not), but not with any of the intermediate structures (semi-group, monoid, group, etc), unlike us. However, their implementation has better performances due to the use of sparse normal form and more optimised algorithms.
Our automatic reflection was written with Idris reflection mechanism which allows to do pattern matching on syntax, and their automatic reflection is programmed in Ltac~\cite{DelahayeLTac} : a proof dedicated and untyped meta-language for the writing of automations. \\

Proof automations have been investigated for various proof assistants and programming languages. In Coq, appart from the ring prover, there's also the omega solver~\cite{Cregut04}, which solves a goal in Presburger arithmetic (i.e. a universally quantified formula made of equations and inequations), and a field~\cite{DelahayeField} decision procedure for real numbers, which plugs to Coq's ring prover after simplification of the multiplicative inverses. Various automations have also been done in Agda~\cite{DBLP:conf/mpc/KokkeS15} ~\cite{Lindblad04}. \\

Proofs by reflection has been intensively studied as in~\cite{ChlipalaBook} and~\cite{Malecha14}, but without anything similar to the type-safe reflection that we have presented here. \\

Finally, if we leave the ground of nice mathematical structures, one can decide to work with arbitrary rewriting rules, but in the general case there isn't a complete decision procedure for such systems, because there is usually no normal form. This is where deduction modulo~\cite{Dowek03} ~\cite{DelahayeModulo} and proof search heuristics starts. \\

>>>>

Coq's ring solver~\cite{Coq2005}, like ours, is
implemented using proof-by-reflection techniques, but without the guarantees obtained with
our type-safe reflection mechanism, and without the correct-by-construction
approach: first, they define the normalisation of terms, then
they prove correctness of the normalisation with an external lemma:
$\forall\ (e1\ e2\ :\ \code{Expr}),\ beq_{Expr}\ (\code{norm}\ e1)\
(\code{norm}\ e2)\ =\ true\ \rightarrow\ \code{reify}\ e1\ \simeq\
\code{reify}\ e2$.
This needs a \code{reify} function that we do not
in our development. Another difference is that Coq's prover only deals
with rings and semi-rings (commutative or not), but not with any of the
intermediate structures (semi-group, monoid, group, etc), unlike our solver. 
However,
their implementation has better performances due to the use of sparse normal
form and more optimised algorithms.  Our automatic reflection was written with
Idris reflection mechanism which allows pattern matching on syntax, and
their automatic reflection is programmed in Ltac~\cite{DelahayeLTac}, a proof
dedicated and untyped meta-language for the writing of automations.
More recently, the Mtac extension~\cite{Ziliani13} provides a
typed language for implementation proof automations.

Proof automations have been investigated for various proof assistants and
programming languages. In Coq, appart from the ring prover, there's also the
Omega solver~\cite{Cregut04}, which solves a goal in Presburger arithmetic
(i.e. a universally quantified formula made of equations and inequations), and
a field~\cite{DelahayeField} decision procedure for real numbers, which plugs
to Coq's ring prover after simplification of the multiplicative inverses.
Various automations have also been done in
Agda~\cite{DBLP:conf/mpc/KokkeS15,Lindblad04}.
%
Proofs by reflection has been intensively
studied~\cite{ChlipalaBook,Malecha14}, but again without anything similar to
the type-safe reflection that we have presented here.

If we leave the ground of nice mathematical structures, one can decide
to work with arbitrary rewriting rules, but in the general case there isn't a
complete decision procedure for such systems, because there is usually no
normal form. This is where deduction modulo~\cite{Dowek03,DelahayeModulo} and
proof search heuristics start.

### 2017-05-20 paper-CICM-17/content/5-conclusions.tex
Add a bit more stuff, prune the bibliography

We have implemented a very generic solution to the initial problem of index mismatch that appeared in section~\ref{sect:motivatingExample} when we were using indexed types. This solution takes the form of a hierarchy of provers for equivalences in algebraic structures. These provers are generic in multiple ways: they work for many algebraic structures (semi-group, monoid, commutative monoid, group, commutative group, ring and semi-ring), for any type that behaves as one of these structures, and for any equivalence relation on this type (and not only for the propositional equality). The implementation is modular and each prover reuses the prover of the structure from which it inherits. These provers can automatically prove equivalences between terms, which discharges the user of providing proofs for some proof obligations, as \code{adc\_lemma\_2}. Thus, these provers enable the user to focus on the interesting proofs that requires specific knowledge and creativity, instead of wasting time for proving routine lemmas that can be automated.\\

The correct-by-construction method that we have followed has involved the design of a type-safe reflection mechanism where reflected terms are indexed over the concrete inputs, and from which we are able to pull out the proofs easily. The construction of the proof is done step by step, and it follows the construction of the normalised terms. Therefore, unlike Coq's and Agda's ring provers, we do not have the traditional duplication that happens when separating the computational content from the proof of correctness. Instead, in our approach, the construction of the proof is guided by the construction of the normalised terms, which, in addition of avoiding redundancy, simplifies the proof generation considerably. This development shows that if dependent types effectively bring some new problems, they are also very expressive tools for building correct-by-construction software where the development is \emph{driven by the types}. \\

>>>>

We have implemented a very generic solution to the initial problem of index
mismatch that appeared in section~\ref{sect:motivatingExample} when we were
using indexed types. This solution takes the form of a hierarchy of provers for
equivalences in algebraic structures. These provers are generic in multiple
ways: they work for many algebraic structures (semi-group, monoid, commutative
monoid, group, commutative group, ring and semi-ring), for any type that
behaves as one of these structures, and for any equivalence relation on this
type (and not only for the propositional equality). The implementation is
modular and each prover reuses the prover of the structure from which it
inherits. These provers can automatically prove equivalences between terms,
which discharges the user of providing proofs for some proof obligations, as
\code{adc\_lemma\_2}. Thus, these provers enable the user to focus on the
interesting proofs that requires specific knowledge and creativity, instead of
wasting time for proving routine lemmas that can be automated.

The correct-by-construction method that we have followed has involved the
design of a type-safe reflection mechanism where reflected terms are indexed
over the concrete inputs, and from which we are able to pull out the proofs
easily. The construction of the proof is done step by step, and it follows the
construction of the normalised terms. Therefore, unlike Coq's and Agda's ring
provers, we do not have the traditional duplication that happens when
separating the computational content from the proof of correctness. Instead, in
our approach, the construction of the proof is guided by the construction of
the normalised terms, which, in addition of avoiding redundancy, simplifies the
proof generation considerably. This development shows that if dependent types
effectively bring some new problems, they are also very expressive tools for
building correct-by-construction software where the development is \emph{driven
by the types}.
